A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-02-12 doc/large_strain.tex
fix closing bracket
update figures and reported in the text 2D average performance

\section{\changeJP{Description of numerical framework}
\label{sec: Description of numerical framework}
}

The discretized system of linear equations formulated in Section \ref{sec:fe}, in conjunction with the constitutive model described in Section \ref{sec: Constitutive modelling} were implemented using \texttt{deal.II} version 9.0 \cite{dealII90}, an open-source finite element library.
This library offers a number of paradigms by which to parallelize a finite-element code.
In this work we employ \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library to perform a standard domain decomposition, whereafter each MPI process ``owns'' only a subset of elements and the mesh is distributed across all MPI processes.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fine_level_2d.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_fine_level}
\end{figure}
%
%
Figure \ref{fig:miehe_fine_level} illustrates a distributed MPI decomposition on the fine mesh level for the 2D problem that will be introduced later in Section \ref{sec:example}.

In practice this implies that parallelization of the assembly operation (for the matrix-based approach) and the linear solver (as well as various pre- and post-processing steps), both of which are the focal points of investigation in this manuscript, has been achieved using MPI.
For the linear solver, \texttt{deal.II}'s implementation of the preconditioned conjugate gradient (CG) solver for symmetric positive-definite systems is consistently used, leaving the choice for method of assembly and linear system preconditioning remaining.
For the matrix-based approach, the Trilinos \cite{Heroux2005} library, version 12.12.1 was leveraged;
the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the package ML package \cite{Gee2006a}.
The second implementation, which is discussed in detail in the following Sections, utilized a matrix-free approach in conjunction with a geometric-multigrid preconditioner.
As will be detailed later, the pre-existing matrix-free framework that was used offers further opportunity for low-level parallelism that may be exploited.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements\footnote{%
\changeJP{In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.}
}.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}

@@

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number $N$ of unknown DoFs is desirable.
The efficiency of standard iterative solvers and preconditioners deteriorates for large systems of linear equations due to increasing iteration counts under mesh refinement.
In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
\changeJP{Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.}

For this work, we adopt the matrix-free level operators within the GMG.
In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial. One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent stiffness tensor is the same.
}.
In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
Essentially, level operators correspond to the linearization around the smoothed representation of displacement field\footnote{
Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
}.

{\color{red}
To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
The efficiency of the preconditioner with the considered level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
% Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
}

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
%
Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{deformed.png}
      \caption{2D deformed mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{deformed_3d.png}
    \caption{3D deformed mesh}
  \end{subfigure}
  \caption{Deformed meshes at the final loading step with quadratic elements and two global mesh refinements.}%
  \label{fig:miehe_deformed}
\end{figure}
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
%
\changeJP{The same distributed approach to mesh decomposition (as most evident on the finest level) is adopted on the multigrid levels.}
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements \changeJP{for the mesh introduced in figure \ref{fig:miehe_fine_level}}.
%
\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
  }
  \label{tab:input_parameters_2d}
\end{table}

@@

\changeJP{%
Aligned with the approach taken in similar previous investigations \cite{}\todo{Cite Martin's papers?}, in this study we limit ourselves to comparisons made on a node level.
Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange through interconnects via MPI.
}
The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
{\color{red}
We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach \changeJP{with a high polynomial order finite element}.
\changeJP{Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound}.
}

%
% Below are are implementational details that might affect the comparative result
%
\changeJP{We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.}
The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level, we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

Computations \changeJP{were performed} on a single node of two clusters:
A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. \changeJP{one} fully occupied node.}

@@

{\color{red}
Figure \ref{fig:roofline} shows results of the roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

>>>>

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

Note that the SIMD vectorization
{\color{red}
in \texttt{deal.II} \cite{dealII90}
}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
{\color{red}
The main reason is that the operations are typically the same on all elements.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, {\color{red}the} deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the \textit{VectorizedArray} templated generic class that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that already supports all the necessary contraction (such as $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$) to implement the three algorithms proposed above.
Thanks to the object oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the so-called locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in deal.II (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and shared memory parallelization as well SIMD vectorization) and performance tests, see \cite{kronbichler12}.
}

@@

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number $N$ of unknown DoFs is desirable.
The efficiency of standard iterative solvers and preconditioners deteriorates for large systems of linear equations due to increasing iteration counts under mesh refinement.
In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.

For this work we adopt the matrix-free level operators within the GMG.
In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial. One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}\footnote{
  We note that this approach is limited to the case where for each quadrature point within an element the fourth-order tangent stiffness tensor is the same.
}.
In this approach, the fine-scale displacement is decomposed into long-wave and short-wave contributions by an additive split, where the former is associated to the homogeneous response of a patch of elements and the latter represents fluctuations due to the heterogeneous structure of the patch.
However, for a patch of elements consisting of the same material, the transfer operators reduce to the standard geometric transfer.
Based on this observation we adopt the following approach: It is assumed that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
Essentially, level operators correspond to the linearization around the smoothed representation of displacement field\footnote{
Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.
}.

{\color{red}
To summarize, in this section we propose a method by which to construct level tangent operators for finite-strain elasticity to be employed within the GMG preconditioner.
The other parts of the GMG preconditioner such as level smoothers, matrix-free interlevel transfer operators and the coarse level iterative solver are provided by the \texttt{deal.II} library.
The efficiency of the preconditioner with the considered level operators will be assessed by numerical examples in the next section.
We will, however, refrain from studying the machine performance aspects of the multigrid preconditioner as a whole as the focus of this paper is the matrix-free implementation of tangent operators of finite-strain elasticity both at the fine mesh level as well as the coarser multigrid levels.
Finally, we note that at the time of writing, a $p-$ version of the GMG was not available within the \texttt{deal.II} library and therefore not considered here.
}

@@

In this section, we apply the proposed matrix-free operator algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner to a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ direction for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
{\color{red}
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fourth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level we use {\color{red}the Chebyshev preconditioner as a solver \cite{Varga2009}} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.}

Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{deformed.png}
      \caption{2D deformed mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{deformed_3d.png}
    \caption{3D deformed mesh}
  \end{subfigure}
  \caption{Deformed meshes at the final loading step with quadratic elements and two global mesh refinements.}%
  \label{fig:miehe_deformed}
\end{figure}

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{fine_level_2d.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D mesh after two global refinements distributed into three MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_fine_level}
\end{figure}

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{gmg_2d_4procs.png}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{2D multigrid mesh after two global refinements distributed into four MPI processes. The color indicates the process owning the element.}%
  \label{fig:miehe_gmg}
\end{figure}
%
%
{\color{red}
MPI parallelization is achieved using the standard domain decomposition, where each MPI process ``owns'' only a subset of elements.
To that end, \texttt{deal.II}'s interface to \texttt{p4est} \cite{p4est} library is used.
Figure \ref{fig:miehe_fine_level} illustrates such decomposition on the fine mesh level for the 2D problem.
The same approach is adopted on the multigrid levels.}
Figure \ref{fig:miehe_gmg} illustrates the hierarchy of multigrid meshes for the 2D problem after two global refinements.
{\color{red}
Implementational details of the MPI-parallel multigrid method in \texttt{deal.II} are discussed in \cite{Clevenger2018}.
}

\begin{table}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{gref}$ & $N_{el}$ & $N_{DoF}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \caption{Parameters for the 2D benchmark: $p$ is the polynomial degree,
  $q$ is the number of quadrature points in 1D, $N_{gref}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{DoF}$ is the number of DoFs.
  }
  \label{tab:input_parameters_2d}
\end{table}

@@

{\color{red}In this study, we limit ourselves to a comparison on a node level, which should already indicate the differences between the matrix-free and matrix-based approaches.}
The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
{\color{red}
We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach.
On the other hand, the problem size is large enough so that the sparse matrix will not fit into the L3 cache and therefore we can observe that the matrix-based approach is memory bound.
}
%
The results of the matrix-free approach with a geometric-multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML of the Trilinos \cite{Heroux2005} library, version 12.12.1.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
Computations are done on a single node of two clusters:
A node on the ``Emmy'' cluster at RRZE, FAU has two Xeon 2660v2 ``Ivy Bridge'' chips (10 cores per chip + SMT) running at 2.2 GHz with 25 MB shared cache per chip and 64 GB of RAM. Intel's compiler version 18.03 with flags ``-O3 -march=native'' and Intel MPI version 18.03 are used.
The ``IWR'' results were obtained on a single machine with eight Xeon E7-8870 ``Sandy Bridge'' chips (10 cores per chip  + SMT) running at 2.4 GHz with 30 MB shared cache per chip. For the simulation, GCC version 8.1.0 with flags ``-O3 -march=native'' and OpenMPI version 3.0.0 were used.
{\color{red}Unless noted otherwise, the simulations are performed with 20 MPI processes, i.e. fully occupied node.}

@@

{\color{red}
Figure \ref{fig:roofline} shows results of the Roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $\pgfmathprintnumber{8.2093245}\,\rm{GFlop/s}$ (4.6\% of the peak performance) in 2D.
The matrix-free implementations of the fine-strain elastic tangent operator (on average) lead to a much better performance of
$\pgfmathprintnumber{18.8047083222}\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory-bound regime.
The achieved performance is about 10.6\% of the peak arithmetic performance. This is much lower than the
70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
Unfortunately, performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.
